---
title: "Assignment_5"
author: "Nick Climaco"
date: "2023-02-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE,tidy.opts=list(width.cutoff=90),tidy=TRUE)
```

### Assignment Description 

The chart above describes arrival delays for two airlines across five destinations. Your task is to:
(1) Create a .CSV file (or optionally, a MySQL database!) that includes all of the information above.
You’re encouraged to use a “wide” structure similar to how the information appears above, so
that you can practice tidying and transformations as described below.
(2) Read the information from your .CSV file into R, and use tidyr and dplyr as needed to tidy
and transform your data.
(3) Perform analysis to compare the arrival delays for the two airlines.
(4) Your code should be in an R Markdown file, posted to rpubs.com, and should include narrative
descriptions of your data cleanup work, analysis, and conclusions. Please include in your
homework submission:
 The URL to the .Rmd file in your GitHub repository. and
 The URL for your rpubs.com web page.

### Import packages
```{r}
library(tidyverse)
```
### Load the Data
```{r}
df <- read.csv("https://raw.githubusercontent.com/Nick-Climaco/Rdataset/main/Numbersenese_kaiserfung_data.csv", na.strings = c("","NA"))
df[2,1] <- "ALASKA"
df[5,1] <- "AM West"
df
```
### Tidy and Clean 

```{r}
df_filter <- df %>%
    filter(rowSums(is.na(.)) < ncol(.)) %>%
    rename(Airline = "X", Status = "X.1")
df_filter
```
### Convert to long format using pivot_longer()
```{r}
df_long <- df_filter %>%
    pivot_longer(cols = c("Los.Angeles", "Phoenix", "San.Diego", "San.Francisco", "Seattle"), names_to = "City", values_to = "Count")

head(df_long)

#cols specifies the columns that will be pivoted 
# names_to specifies the name of the column that will hold the name/s of the old column name/s
#values_to specifies the name of the column that will hold the values of the old column/s
```
### Analysis : for arrival delays for Alaska and AM West airlines

```{r}
df_long %>% 
    filter(Airline == "ALA") %>% 
    summarise(
        n_delays = sum(Count[Status == "delayed"]),
        total_flights = sum(Count),
        delay_rate = round(n_delays / total_flights ,2)
    )
```

```{r}
df_summary <- df_long %>% 
    group_by(Airline, City) %>%
    summarise(
        total_delays = sum(Count[Status == "delayed"]),
        total_flights = sum(Count),
        pct_delays = round(total_delays / total_flights, 2)
    )
df_summary  
```
### Visualization
```{r}
ggplot(df_summary, aes(x = City, 
                       y = pct_delays, 
                       fill = Airline)) +
  geom_bar(stat = "identity", 
           position = "dodge") +
  geom_text(aes(label = pct_delays), 
            position = position_dodge(width = 1),
            vjust = -0.5) +
  labs(x = "Airline", 
       y = "Delay rate", 
       fill = "Airline") +
  theme_bw()

```

### Conclusion

From the given data and shown on the graph, it can be observed that AM West Airlines had a higher rate of arrival delays in these five cities relative to Alaska Airlines. The difference in delays rate is at least 3% which is indicative of a performance disparity between the two airlines in this closed experiment. Thus, the data suggests that traveling to these five cities, Alaska Airlines might be the better choice since it has a lower chance of a delayed arrival. 

